---
title: "Alpha Diversity and Beta Dispersion Analyses and Figures"
author: "Karolina Zabinski"
date: "2024-07-03"
output: pdf_document
---

# Introduction:  
The following script is to analyse alpha diversity and beta dispersion found in SM Fig. 5A-D and ASV richness analyses corresponding to Fig. 2B.  

```{r, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(width.cutoff = 60), root.dir ="~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment" )
```  

Let's load in our libraries
```{r, message = FALSE}
library(dplyr) #manipulate df
library(tidyverse)
library(doBy)
library(gridExtra)
library(ggpubr)
library(formatR)
library(reshape2)
library(tidyr)
library(ggplot2)
library(lme4)
library(car)
library(emmeans)
library(gridExtra)
library(cowplot)
library(vegan)
library(usethis)
library(phyloseq)
library(speedyseq)
```  
# General themes and equations  

```{r, echo=TRUE, results='hide'}
# set horizontal offset for points so they are easier to see
hdodge <- position_dodge(width=0.1) 

# set up theme:
themes <-  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

# loading in function
fun <- function(x,...) {c(m=mean(x, na.rm=T), sd=sd(x,na.rm=T), n = length(x), se=sd(x,na.rm=T)/sqrt(length(x)))}

# create distinct color palette for populations
colors.vec <- c("black", "white")

# shapes for the two levels of PlantOrigin
shapes.vec <- c(16, 21)  
  # 16 for black dot & 21 for black dot with white fill
```
# Alpha diversity, SM Fig 5A-C:  

Load in filtered ps object.  
```{r}
singletons.ps <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/ps.no0.rds")
singletons.ps %>% otu_table %>% print(n=5) # looks good
length(which(taxa_sums(singletons.ps) == 1))
```  
Create sample data frame.  
```{r}
Sing.sample.df <- data.frame(sample_data(singletons.ps))
str(Sing.sample.df) # looks good
```  
Estimate alpha diversity metrics.  
```{r, echo=F}

# estimate alpha diversity (non-rarefied)
Sing.rich <- estimate_richness(singletons.ps, measures=c("Observed", "Shannon", "Simpson", "InvSimpson")) # 

#make into a dataframe
Sing.alphdiv <- as.data.frame(Sing.rich)

#For merging, name ID columns the same thing
Sing.alphdiv$sample.no <- rownames(Sing.alphdiv)

# merge alpha diversity and sample meta-data
Sing.alpha <- merge(Sing.alphdiv, Sing.sample.df, by = "sample.no")

# add column for Hill number order 1 (q=1): 
Sing.alpha$Hill_q1 <- exp(Sing.alpha$Shannon)

# Hill no. order 2 (q=2) is already calculated (InvSimpson column)

# save as RDS
saveRDS(Sing.alpha, file = "~/Box Sync/GSE/GrainSizeExperiment/output/singletons.alphaDiv.rds" )

# loading in rds
Sing.alpha <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/singletons.alphaDiv.rds")
print(Sing.alpha)
```  
## Shannon diversity Fig 5B:  
Visualization:  
```{r}
# subset df
singletons.shannon.df <- Sing.alpha %>% select(plant.origin, sed.origin, tank.id, Shannon)

# calculate summary stats
singletons.shannon.sum <- summaryBy(Shannon ~ plant.origin + sed.origin, data=singletons.shannon.df, FUN= fun)

alpha.sig.label1 <- paste("Pop, p = 0.10")
alpha.sig.label2 <- paste("Sed, p = 0.49")
alpha.sig.label3 <- paste("PopxSed, p = 0.42")

# plot
Sing.shannon.plot <- ggplot(singletons.shannon.sum, aes(x=sed.origin, y=Shannon.m, group=plant.origin, color=plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = Shannon.m + Shannon.se, ymin = Shannon.m - Shannon.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0), limits = c(4.5, 5.5)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  themes + 
  scale_x_discrete(labels= c("Blake's\nLanding","Millerton\nPoint")) +
  labs(y = expression("Shannon diversity"), x= expression("Sediment Origin")) +
  theme( legend.position = c(0.86, 0.85),plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 20)) +
  annotate("text", x = 1.9, y = 5.46, label = alpha.sig.label1, size = 3) +
  annotate("text", x = 1.9, y = 5.4, label = alpha.sig.label2, size = 3) +
  annotate("text", x = 1.9, y = 5.34, label = alpha.sig.label3, size = 3)

Sing.shannon.plot
```  
Analysis.  
```{r}
# model selection
shannon.2way <- lmer(Shannon ~ plant.origin*sed.origin + plant.origin + sed.origin + (1|tank.id), data=Sing.alpha, na.action = na.exclude)
plot(shannon.2way)
qqnorm(resid(shannon.2way))
hist(resid(shannon.2way)) # normal enough
shapiro.test(resid(shannon.2way))
summary(shannon.2way) # passes
Anova(shannon.2way, type= 2) 
```  
## Hill q=1 number SM Fig 5B:  
```{r}
# subset df
singletons.Hill1.df <- Sing.alpha %>% select(plant.origin, sed.origin, tank.id, Hill_q1)

# calculate summary stats
singletons.Hillq1.sum <- summaryBy(Hill_q1 ~ plant.origin + sed.origin, data=singletons.Hill1.df, FUN= fun)

hill.sig.label1 <- paste("Pop, p = 0.10")
hill.sig.label2 <- paste("Sed, p = 0.44")
hill.sig.label3 <- paste("PopxSed, p = 0.26")

# plot
Sing.Hillq1.plot <- ggplot(singletons.Hillq1.sum, aes(x=sed.origin, y=Hill_q1.m, group=plant.origin, color=plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = Hill_q1.m + Hill_q1.se, ymin = Hill_q1.m - Hill_q1.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0), limits = c(115, 265)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec,labels= c("Blake's\nLanding","Millerton\nPoint")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  themes + 
  scale_x_discrete(labels= c("Blake's\nLanding","Millerton\nPoint")) +
  labs(y = expression("Hill no. 1"), x= expression("Sediment Origin")) +
  theme( legend.position = c(0.86, 0.85),plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 20)) +
    annotate("text", x = 1.9, y = 260, label = hill.sig.label1, size = 3) +
  annotate("text", x = 1.9, y = 252, label = hill.sig.label2, size = 3) +
  annotate("text", x = 1.9, y = 244, label = hill.sig.label3, size = 3)
Sing.Hillq1.plot
```  
Analysis:  
```{r}
# model selection
hill.2way <- lmer(Hill_q1 ~ plant.origin*sed.origin + plant.origin + sed.origin + (1|tank.id), data=singletons.Hill1.df, na.action = na.exclude)
plot(hill.2way)
qqnorm(resid(hill.2way))
hist(resid(hill.2way)) 
shapiro.test(resid(hill.2way)) # pass
summary(hill.2way) 
Anova(hill.2way, type= 2) 
```  
## Richness Fig 2B:  
```{r}
# subset df
singletons.rich.df <- Sing.alpha %>% select(plant.origin, sed.origin, tank.id, Observed)

# calculate summary stats
singletons.rich.sum <- summaryBy(Observed ~ plant.origin + sed.origin, data=singletons.rich.df, FUN= fun)

# plot data
Sing.rich.plot <- ggplot(singletons.rich.sum, aes(x=sed.origin, y=Observed.m, group=plant.origin, color=plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = Observed.m + Observed.se, ymin = Observed.m - Observed.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0), limits =c(2100,2950)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  themes + 
  scale_x_discrete(labels= c("Blake's\nLanding","Millerton\nPoint")) +
  labs(y = expression("ASV Richness"), x= expression("Sediment Origin")) +
  theme( legend.position = c(0.86, 0.85),plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 20))
Sing.rich.plot
```  
Analysis:  
```{r}
# model selection
rich.2way <- lmer(Observed ~ plant.origin*sed.origin + plant.origin + sed.origin + (1|tank.id), data=singletons.rich.df, na.action = na.exclude)
plot(rich.2way)
qqnorm(resid(rich.2way))
hist(resid(rich.2way)) # normal enough
shapiro.test(resid(rich.2way)) # pass
summary(rich.2way) 
Anova(rich.2way, type= 2) 
```  
## Inverse Simpson SM Fig 5C:  
```{r}
# subset df
singletons.invSimp.df <- Sing.alpha %>% select(plant.origin, sed.origin, tank.id, InvSimpson)

# calculate summary stats
singletons.invSimp.sum <- summaryBy(InvSimpson ~ plant.origin + sed.origin, data=singletons.invSimp.df, FUN= fun)

invSimp.sig.label1 <- paste("Pop, p = 0.15")
invSimp.sig.label2 <- paste("Sed, p = 0.38")
invSimp.sig.label3 <- paste("PopxSed, p = 0.51")

# plot
Sing.invSimp.plot <- ggplot(singletons.invSimp.sum, aes(x=sed.origin, y=InvSimpson.m, group=plant.origin, color=plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = InvSimpson.m + InvSimpson.se, ymin = InvSimpson.m - InvSimpson.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0), limits = c(20,55)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  themes + 
  scale_x_discrete(labels= c("Blake's\nLanding","Millerton\nPoint")) +
  labs(y = expression("Inverse Simpson"), x= expression("Sediment Origin")) +
  theme( legend.position = c(0.86, 0.85),plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 20)) +
  annotate("text", x = 1.9, y = 25, label = invSimp.sig.label1, size = 3) +
  annotate("text", x = 1.9, y = 23, label = invSimp.sig.label2, size = 3) +
  annotate("text", x = 1.9, y = 21, label = invSimp.sig.label3, size = 3)
Sing.invSimp.plot
```  
Analysis:  
```{r}
# model selection
invSimp.2way <- lmer(InvSimpson ~ plant.origin*sed.origin + plant.origin + sed.origin + (1|tank.id), data=singletons.invSimp.df, na.action = na.exclude)
plot(invSimp.2way)
qqnorm(resid(invSimp.2way))
hist(resid(invSimp.2way)) 
shapiro.test(resid(invSimp.2way)) # pass
summary(invSimp.2way) 
Anova(invSimp.2way, type= 2) 
```  
# Beta Dispersion SM Fig 5D:  
Loading in relative abundance ps object.  
```{r}
rel.ps <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/ps_rel.rds")
rel.ps %>% otu_table %>% print(n=5) # looks good
```  
Calculating dispersion for analyses across population and sediment origin, then saving as RDS.   
```{r}
# make your otu_table into a matrix
otu.rel.matrix <- as.matrix(otu_table(rel.ps)) 

# isolate metadata
metadata.df <- sample_data(rel.ps)

# Groups being compared: plant.origin & sed.origin

# make new column variable combining other factors
metadata.df$combined_trt <- paste(metadata.df$plant.origin, metadata.df$sed.origin, sep= "")

# get Bray Curtis distance between samples 
all.bc <- vegan::vegdist(otu.rel.matrix, method="bray") 

# calculate multivariate dispersion
# dispersion of samples relative to group median 
# for each plant origin x sediment origin type combo
all.d <-betadisper(all.bc, group=metadata.df$combined_trt) 

# test for differences in dispersions between groups
anova(all.d) 

# combine dispersion data with meta data
all.disp.df <- as.data.frame(all.d$distances)
all.disp.df$Sample.id <- rownames(all.disp.df)
colnames(all.disp.df)[1] <- "Dispersion"
metadata.df <- data.frame(metadata.df)
metadata.df$Sample.id <- rownames(metadata.df)

# merge dispersions with metadata
all.disp.df2 <- merge(all.disp.df, metadata.df, by = "Sample.id")

# save as RDS
saveRDS(all.disp.df2, file = "~/Box Sync/GSE/GrainSizeExperiment/output/dispersion.rds")
```  
Visualizations. 
```{r}
# calculate summary stats
disp.sum <- summaryBy(Dispersion ~ plant.origin + sed.origin, data=all.disp.df2, FUN= fun)

disp.sig.label1 <- paste("Pop, p = 0.05")
disp.sig.label2 <- paste("Sed, p = 0.44")
disp.sig.label3 <- paste("PopxSed, p = 0.98")

# plot data
disp.plot <- ggplot(disp.sum, aes(x=sed.origin, y=Dispersion.m, group=plant.origin, color=plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = Dispersion.m + Dispersion.se, ymin = Dispersion.m - Dispersion.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0.25, 0.50)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  themes + 
  scale_x_discrete(labels= c("Blake's\nLanding","Millerton\nPoint")) +
  labs(y = expression("Beta Dispersion"), x= expression("Sediment Origin")) +
  theme( legend.position = c(0.86, 0.85),plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 20)) +
    annotate("text", x = 1.9, y = 0.29, label = disp.sig.label1, size = 3) +
  annotate("text", x = 1.9, y = 0.275, label = disp.sig.label2, size = 3) +
  annotate("text", x = 1.9, y = 0.26, label = disp.sig.label3, size = 3)
disp.plot
```  
Analysis:  
```{r}
# model selection
disp.2way <- lmer(Dispersion ~ plant.origin*sed.origin + plant.origin + sed.origin + (1|tank.id), data=all.disp.df2, na.action = na.exclude)
plot(disp.2way)
qqnorm(resid(disp.2way))
hist(resid(disp.2way)) 
shapiro.test(resid(disp.2way)) # passes
summary(disp.2way) 
Anova(disp.2way, type= 2) 
```  
# Grouping alpha diversity figures:  
```{r}
alphaTraits.plot <- ggarrange(Sing.shannon.plot, Sing.Hillq1.plot,Sing.invSimp.plot,disp.plot,
                    labels = c("a.", "b.", "c.", "d."), ncol=2, nrow=2, common.legend = TRUE, legend="bottom")
alphaTraits.plot

ggsave(filename = "~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/plant-figures/SI-fig3.png", width=8, height =8, alphaTraits.plot)
```  