---
title: "Community and Distances Analyses and Figures"
author: "Karolina Zabinski"
date: "2024-07-03"
output: html_document
---
# Introduction:  
This script is for community analyses and distances found in figures 4A-C.  

```{r, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(width.cutoff = 60), root.dir ="~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment" )
```  

Let's load in our libraries
```{r, message = FALSE}
library(dplyr) #manipulate df
library(tidyverse)
library(doBy)
library(gridExtra)
library(ggpubr)
library(formatR)
library(reshape2)
library(tidyr)
library(ggplot2)
library(lme4)
library(car)
library(emmeans)
library(gridExtra)
library(cowplot)
library(vegan)
library(usethis)
library(phyloseq)
library(speedyseq)
library(permute)
```  
# General themes and equations:   
```{r, echo=TRUE, results='hide'}
# set horizontal offset for points so they are easier to see
hdodge <- position_dodge(width=0.7) 

# set up theme:
themes <-  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

# loading in function
fun <- function(x,...) {c(m=mean(x, na.rm=T), sd=sd(x,na.rm=T), n = length(x), se=sd(x,na.rm=T)/sqrt(length(x)))}
```  
# Only S Cycling Families, Fig 5A:  
First, I want to see how an nmds of all populations compare when I only include these DA families.  
```{r}
# load data
rel.ps <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/ps_rel.rds")

# create vec of fam of interest
fam.vec <- c( "Desulfatiglandaceae", "Desulfobacteraceae", "Desulfobulbaceae", "Desulfocapsaceae","Desulfosarcinaceae","Desulfovibrionaceae","Desulfuromonadaceae","Dethiosulfatibacteraceae","Sulfurimonadaceae","Sulfurovaceae","Thioalkalispiraceae","Thiomicrospiraceae")

# subset rel.ps based on fam.vec
DA.fam.ps <- subset_taxa(rel.ps, Family %in% fam.vec)  %>% 
            filter_taxa( function(x) sum(x) > 0, TRUE)
```  
Then, I want to visualize this data in an nmds to determine whether sediment or plant origin drive microbiome assembly at the ASV level.  
```{r, echo=TRUE, message=FALSE, results='hide'}
# make ordination plots with weighted Bray-Curtis dissimilarity matrix 
DAFam.ord.nmds.bray <- ordinate(DA.fam.ps,  method="NMDS", distance="bray", weighted=TRUE, trymax = 5000, strata=tank.id) 
```  
Checking BC ordination stats.   
```{r}
# check stress to make sure 2 dimensions are adequate
  # (need more if stress ~ >.2)
DAFam.ord.nmds.bray$stress # good
  # 0.09832771

# Shepards test of goodness of fit
goodness(DAFam.ord.nmds.bray) 
  # Lists test statistics of goodness of fit for each point

stressplot(DAFam.ord.nmds.bray) # looks good
  # Produces a Shepards diagram
```   
Plot ordination.  
```{r}
# create distinct color palette for populations
colors.vec <- c("darkgoldenrod", "orchid4")

# shapes for the two levels of PlantOrigin
shapes.vec <- c(16, 21)  
  # 16 for black dot & 21 for black dot with white fill

DAFam.sig.label1 <- paste("Pop, p = 0.31")
DAFam.sig.label2 <- paste("Sed, p = 0.00")

DAFam.nmds.plot <- plot_ordination(DA.fam.ps, DAFam.ord.nmds.bray, color="sed.origin", shape= "plant.origin") +
  geom_point(aes(shape = plant.origin), size = 5, alpha=.7) + themes +
  scale_fill_manual(name = "Sediment Origin", values = colors.vec, labels= c("Blake's\nLanding", "Millerton\nPoint")) +
  scale_shape_manual(name = "Population Origin", values = c(16,21), labels= c("Blake's\nLanding", "Millerton\nPoint")) +  # Set shapes manually
  scale_color_manual(name = "Sediment Origin", values = colors.vec, labels= c("Blake's\nLanding", "Millerton\nPoint")) +
  annotate("text", x = -.5, y = .9, label = "S Cyclers", size = 6) +
  guides(shape = guide_legend(title = "Populatioin Origin"))+ 
  theme(text = element_text(size = 20)) +
  annotate("text", x = -0.5, y = .8, label = DAFam.sig.label1, size = 6) +
  annotate("text", x = -0.5, y = .7 , label = DAFam.sig.label2, size = 6) +
  annotate("text", x = 0.8, y = -.8, label = "stress = 0.10", size = 6)
DAFam.nmds.plot
```  
Okay so with only the DA families related to sulfur processing, we do not see any difference in population so it's not those related to sulfur processing. Maybe it's taxa not associated with the sulfur cycle. Let's remove those DA taxa and re-run the nMDS.  

Let's run a dbRDA which takes into account randome effects to double check.  
```{r}
# create sample df of all treatments
all.sample.df <- data.frame(sample_data(DA.fam.ps))

# create response df of relative abundances
DA.fam.responses.df <- data.frame(otu_table(DA.fam.ps))

# set up permutation structure that restricts permutations within each tank (random effects)
DA.fam.perm.schem <- how(nperm=999, blocks = all.sample.df$tank.id)

# Setting up dbrda. Avoiding negative eigenvalues with sqrt distances
mod.DA.fam <- dbrda(DA.fam.responses.df ~ plant.origin*sed.origin + plant.origin + sed.origin + Condition(tank.id), all.sample.df, dist="bray", sqrt.dist = TRUE)

# running anova
anova(mod.DA.fam, permutations= DA.fam.perm.schem, by = "margin")
  # two-way interaction is not significant, dropping

mod.DA.fam <- dbrda(DA.fam.responses.df ~ plant.origin + sed.origin + Condition(tank.id), all.sample.df, dist="bray", sqrt.dist = TRUE)

# running anova
anova(mod.DA.fam, permutations= DA.fam.perm.schem, by = "margin")
  # just sediment is significant

# Get adjusted R^2 for the overall model
RsquareAdj(mod.DA.fam)

# Calculating total SumOfSquares
total_SS <- sum(c(0.3000, 1.6827, 4.1513))

# Caluclate R^2 for each term
r2.plant <- 0.300/total_SS
  # 0.0489077
r2.sed <- 1.6827/total_SS
  # 0.2743234
```
Okay so only sediment drives microbiome assembly when only ASVs from DA S cyclers are included.  
# No S Cycling taxa on microbiome assembly, Fig. 5B:  
```{r}
# subset rel.ps excluding families in fam.vec
DA.NoFam.ps <- subset_taxa(rel.ps, !(Family %in% fam.vec))  %>% 
            filter_taxa( function(x) sum(x) > 0, TRUE)
# make ordination plots with weighted Bray-Curtis dissimilarity matrix 
DANoFam.ord.nmds.bray <- ordinate(DA.NoFam.ps,  method="NMDS", distance="bray", weighted=TRUE, trymax = 5000, strata=tank.id) 
```  
Checking BC ordination stats.  
```{r}
# check stress to make sure 2 dimensions are adequate
  # (need more if stress ~ >.2)
DANoFam.ord.nmds.bray$stress # good
  # 0.1606387

# Shepards test of goodness of fit
goodness(DANoFam.ord.nmds.bray) 
  # Lists test statistics of goodness of fit for each point

stressplot(DANoFam.ord.nmds.bray) # looks good
  # Produces a Shepards diagram
```   
Plot ordination.  
```{r}
# create distinct color palette for populations
colors.vec <- c("darkgoldenrod", "orchid4")

# shapes for the two levels of PlantOrigin
shapes.vec <- c(16, 21)  

allFam.sig.label1 <- paste("Pop, p = 0.01")
allFam.sig.label2 <- paste("Sed, p = 0.00")

# plot with colors for each treatment
DANoFam.nmds.plot <- plot_ordination(DA.NoFam.ps, DANoFam.ord.nmds.bray, color="sed.origin", shape= "plant.origin") +
  geom_point(aes(shape = plant.origin), size = 6, alpha=.7) + themes +
  scale_fill_manual(name = "Sediment Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  scale_shape_manual(name = "Population Origin", values = c(16,21), labels= c("Blake's\nLanding","Millerton\nPoint")) +  # Set shapes manually
  scale_color_manual(name = "Sediment Origin", values = colors.vec, labels= c("Blake's\nLanding","Millerton\nPoint")) +
  annotate("text", x = 0.5, y = 1.0, label = "No S Cyclers", size = 6) +
  guides(shape = guide_legend(title = "Population Origin"))+ 
  theme(text = element_text(size = 20)) +
  annotate("text", x = 0.5, y = .9, label = allFam.sig.label1, size = 6) +
  annotate("text", x = 0.5, y = .8 , label = allFam.sig.label2, size = 6) +
  annotate("text", x = 0.6, y = -.8, label = "stress = 0.16", size = 6)
DANoFam.nmds.plot
```  
Okay so it looks like the differences between populations are driven by taxa not associated with the sulfur cycle.

Running dbRDA analysis to include random effects.  
```{r}
# create sample df of all treatments
all.sample.df2 <- data.frame(sample_data(DA.NoFam.ps))

# create response df of relative abundances
no.DA.fam.responses.df <- data.frame(otu_table(DA.NoFam.ps))

# set up permutation structure that restricts permutations within each tank (random effects)
no.DA.fam.perm.schem <- how(nperm=999, blocks = all.sample.df2$tank.id)

# Setting up dbrda. Avoiding negative eigenvalues with sqrt distances
mod.no.DA.fam <- dbrda(no.DA.fam.responses.df ~ plant.origin*sed.origin + plant.origin + sed.origin + Condition(tank.id), all.sample.df2, dist="bray", sqrt.dist = TRUE)

# running anova
anova(mod.no.DA.fam, permutations= no.DA.fam.perm.schem, by = "margin")
  # two-way interaction is not significant, dropping

mod.no.DA.fam <- dbrda(no.DA.fam.responses.df ~ plant.origin + sed.origin + Condition(tank.id), all.sample.df2, dist="bray", sqrt.dist = TRUE)

# running anova
anova(mod.no.DA.fam, permutations= no.DA.fam.perm.schem, by = "margin")
  # just sediment and plant origin are sig

# Calculating total SumOfSquares
total_SS <- sum(c(0.5283, 0.9687, 4.2001))

# Caluclate R^2 for each term
r2.plant <- 0.5283/total_SS
  # 0.0927313896543856
r2.sed <- 0.9687/total_SS
  # 0.170033876884731
```  
Plant and sediment drive differences in microbiome assembly in taxa outside of the sulfur cycling.  

# Pop distances between sediment origin, Fig C:  
## Data manipulation:  
Load in filtered relative abundance  ps object.  
```{r}
rel.ps <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/ps_rel.rds")
```  
Calculate BC dissimilarity matrix and isolate all rows corresponding to MP plants and all cols corresponding to MP plants.  
```{r, echo=TRUE, message=FALSE, results='hide'}
# calculate BC dissimilarity matrix
all.bc <- as.matrix(distance(rel.ps, method = "bray"))

# create vector of names to match rows
  # these are all ids of MP plants
row.names <- c("KZ.4a","KZ.7a","KZ.9a","KZ.15a","KZ.19a","KZ.20a","KZ.6a","KZ.8a","KZ.11a","KZ.12a","KZ.18a","KZ.23a")
col.names <- c("KZ.4a","KZ.7a","KZ.9a","KZ.15a","KZ.19a","KZ.20a","KZ.6a","KZ.8a","KZ.11a","KZ.12a","KZ.18a","KZ.23a")

# create new dissimilarity matrix with only matching rows
subset.distances.bc <- all.bc[row.names, col.names]
  # this is 1/4th of number of entries of original matrix
  # so this works
```  
Now, we need to subset again based on mismatching sediment origin. Isolate all rows corresponding to BL sediment and columns based on MP sediment.  
```{r}
# sediment ids
row.names2 <- c("KZ.4a","KZ.7a","KZ.9a","KZ.15a","KZ.19a","KZ.20a")
col.names2 <- c("KZ.6a","KZ.8a","KZ.11a","KZ.12a","KZ.18a","KZ.23a")

# create new dissimilarity matrix comparing sediment type
subset.distances.MP <- subset.distances.bc[row.names2, col.names2]

# isolate entries and create a new df
mp.distances.df <- data.frame(as.vector((subset.distances.MP)))

# create column names: distance, population
names(mp.distances.df)[1] <- "distances"
mp.distances.df <- mp.distances.df %>% mutate(population = "MILL")
str(mp.distances.df) # looks good!
```  
Now, let's repeat this for BL.  

Calculate BC dissimilarity matrix and isolate all rows corresponding to BL plants and all cols corresponding to BL plants.  
```{r, echo=TRUE, message=FALSE, results='hide'}
# calculate BC dissimilarity matrix
all.bc <- as.matrix(distance(rel.ps, method = "bray"))

# create vector of names to match rows
  # these are all ids of MP plants
row.names <- c("KZ.1a","KZ.3a","KZ.16a","KZ.22a","KZ.24a","KZ.25a","KZ.2a","KZ.5a","KZ.10a","KZ.14a","KZ.17a","KZ.21a")
col.names <- c("KZ.1a","KZ.3a","KZ.16a","KZ.22a","KZ.24a","KZ.25a","KZ.2a","KZ.5a","KZ.10a","KZ.14a","KZ.17a","KZ.21a")

# create new dissimilarity matrix with only matching rows
subset.distances.bc <- all.bc[row.names, col.names]
  # this is 1/4th of number of entries of original matrix
  # so this works
```  
Now, we need to subset again based on mismatching sediment origin. Isolate all rows corresponding to BL sediment and columns based on MP sediment.  
```{r}
# sediment ids
row.names2 <- c("KZ.1a","KZ.3a","KZ.16a","KZ.22a","KZ.24a","KZ.25a")
col.names2 <- c("KZ.2a","KZ.5a","KZ.10a","KZ.14a","KZ.17a","KZ.21a")

# create new dissimilarity matrix comparing sediment type
subset.distances.BL <- subset.distances.bc[row.names2, col.names2]

# isolate entries and create a new df
bl.distances.df <- data.frame(as.vector((subset.distances.BL)))

# create column names: distance, population
names(bl.distances.df)[1] <- "distances"
bl.distances.df <- bl.distances.df %>% mutate(population = "BLAK")
str(bl.distances.df) # looks good!
```  
Now, let's stack the two dfs on top of each other and plot.  
```{r}
# merge dfs
distances.df <- dplyr::bind_rows(bl.distances.df, mp.distances.df)

# save as rds
saveRDS(distances.df, file = "~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/distances.rds")

distances.df <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/distances.rds")
```  
And plot:  
```{r}
# calculate stats for: period, genotype, treatment, family
distances.sum <- summaryBy(distances ~ population, data=distances.df, FUN= fun)

# create distinct color palette for populations
colors.vec <- c("black", "white")

# shapes for the two levels of PlantOrigin
shapes.vec <- c(16, 21) 

dist.sig.label1 <- expression("Pop, p = 0.00")

#plot
distances.plot <- ggplot(distances.sum, aes(x=population, y=distances.m)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = distances.m + distances.se, ymin = distances.m - distances.se), width = 0, position = hdodge, color = "black") + 
  geom_point( color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0)) + 
  themes + 
  scale_x_discrete(labels = c("Blake's\nLanding", "Millerton\nPoint")) +
  labs(y = expression("Bray-Curtis Pairwise Distance"), x= expression("Plant Origin")) +
  theme( legend.position = c(0.86, 0.85),plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 20)) + 
  annotate("text", x = 1.9, y = 0.815, label = dist.sig.label1, size = 6)
distances.plot
```  
Analysis:  
```{r}
# running model  
dist.1way <- lm(distances ~ population, data=distances.df, na.action = na.exclude)
plot(dist.1way)
qqnorm(resid(dist.1way))
hist(resid(dist.1way)) # normal
summary(dist.1way) 
Anova(dist.1way, type= 2) 
```  
There is a difference between these two distances where MP individuals are more similar to each other across sediment types than BL individuals. 

# Group plot of nMDS and distances:  
```{r}
nmdsDistance.plot <- ggarrange(DAFam.nmds.plot, DANoFam.nmds.plot,distances.plot,
                    labels = c("A.", "B.", "C."), ncol=3, nrow=1, common.legend = TRUE, legend="bottom")
nmdsDistance.plot

ggsave(filename = "~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/plant-figures/nmds-distance.png", width=15, height =6, nmdsDistance.plot)
```