---
title: "DAA"
author: "Karolina Zabinski"
date: "2024-07-02"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(width.cutoff = 60), root.dir ="~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment" )
```  

# Introduction:  
This code is for the DAA between sediment treatments (SM Fig 4), and LMM of how S cycling families change across sediment and population origin (Fig. 5, SM Fig. 5, SM Fig. 6)

# Setting up workspace:  
Let's load in our libraries
```{r, message = FALSE}
library(dplyr) #manipulate df
library(tidyverse)
library(doBy)
library(gridExtra)
library(ggpubr)
library(formatR)
library(reshape2)
library(tidyr)
library(ggplot2)
library(lme4)
library(car)
library(emmeans)
library(gridExtra)
library(cowplot)
library(vegan)
library(usethis)
library(phyloseq)
library(speedyseq)
library(corncob)
```  

# General themes and equations  

```{r, echo=TRUE, results='hide'}
# set horizontal offset for points so they are easier to see
hdodge <- position_dodge(width=0.1) 

# set up theme:
themes <-  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

# loading in function
fun <- function(x,...) {c(m=mean(x, na.rm=T), sd=sd(x,na.rm=T), n = length(x), se=sd(x,na.rm=T)/sqrt(length(x)))}

# create distinct color palette for populations
colors.vec <- c("black", "white")

# shapes for the two levels of PlantOrigin
shapes.vec <- c(16, 21)  
  # 16 for black dot & 21 for black dot with white fill
```  
# Introduction:  
Based on the NMDS and PERMANOVA, variation in community assembly is explained by differences at the ASV-level between treatments. Therefore, I will test for differential abundant ASVs between sediment type using beta-binomial regression models (Martin et al. 2019). Then, I will isolate those taxa and see how their relative abundance differs across populations.   

# DAA between Sediment Origin:  
Preparing data.  
```{r}
# load data
all.ps <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/ps.no0.rds")
```  
Run model.  
```{r}
# set reference level to BL sediment
sample_data(all.ps)$sed.origin <- relevel(as.factor(get_variable(all.ps, "sed.origin")), ref="BLAK")

# sediment origin model
set.seed(1)
da_analysis.sed <- differentialTest(
      formula = ~  sed.origin + plant.origin, # full model 
      phi.formula = ~ 1, # no diff in disp b/w groups
      formula_null = ~  plant.origin, # testing effect of sed
      phi.formula_null = ~ 1, # specifying RF for dispersion
      test = "Wald", boot = FALSE,
      data = all.ps,
      fdr_cutoff = 0.05)
#  significant taxa
summary(da_analysis.sed)
  # 943 significant taxa

# extract data and CI from plots
all.efx<- plot(da_analysis.sed, data_only = TRUE, level = c("Order","Family", "Genus"))

# Create a new column 'ASV' to store the text within parentheses
all.efx$ASV <- sub(".*\\(([^)]+)\\).*", "\\1", all.efx$taxa)

# Remove the text within parentheses from the 'taxa' column
all.efx$taxa <- sub("\\s*\\(.*\\)", "", all.efx$taxa)

# change underscores to spaces and commas
all.efx$taxa2<- gsub("_", ", ", all.efx$taxa)

# separate each taxonomic level into it's own col
all.efx <- separate(all.efx, taxa, into = c("Order", "Family", "Genus"), sep = "_")
all.efx$FamGen <- paste(all.efx$Family, all.efx$Genus)
# save output
saveRDS(all.efx,file= "~/Box Sync/GSE/GrainSizeExperiment/output/sed-DA-ASVs.rds")
```  
Plotting:  
```{r}
all.efx.plot <- ggplot(all.efx, aes(x = FamGen, y = x, fill = Family)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "dark grey") +
  geom_errorbar(aes(ymax = xmax, ymin = xmin), width = 0, position = position_dodge(width = 0.2)) +
  geom_point(shape = 21, size = 2.2, alpha = 0.7, position = position_dodge(width = 0.2)) +
  scale_y_continuous(limits = c(-7, 7)) +
  scale_x_discrete(limits = rev) + 
  themes +  
  #scale_fill_manual(values=fam.colors) +
  theme(legend.position = "none") +
  coord_flip() +
  labs(x = "Microbial taxa", y = "Difference in log fold abundance") +
  ggtitle("Differentially abundant taxa from BL to MP sediment") 
all.efx.plot
```  
That's a lot of taxa! I am interested in S-cycling families because sulfur cyclers are hypothesized to be important to maintain in seagrasses for their health. So let's isoalte all sulfur cycling families known to associate with seagrasses + with their functions known/hypothesized. 

Taxa and why:  
Desulfobacteraceae: sulfate reducers (produce sulfide)  
Colwelliaceae: associated with SWD (may not include for this study)  
Desulfatiglandaceae: reduces sulfate into sulfide/strict anaerobe (not found in seagrass literture)  
Desulfobacteraceae: sulfate reducer (produces sulfide)  
Desulfobulbaceae: sulfate reducer 
Desulfocapsaceae: sulfate reducer in anoxic sediments  
Desulfosarcinaceae: sulfate reducer  
Desulfovibrionaceae: strict anaerobe, sulfate reducer  
Desulfuromonadaceae: sulfate reducer  
Dethiosulfatibacteraceae: reduce thiosulfate and elemental sulfur to sulfide.  
Ectothiorhodospiraceae: purple sulfur bacteria, but unclear if sulfate oxidizer or reducer so removing  
Methylomonadaceae: methanotrophic in sulfide rich marine sediments, but won't keep in analysis because I'm focusing on sulfur pathway and this is more bac-bac interactions, and it was 1 ASV that was DA.
Sulfurimonadaceae: sulfide oxidizer  
Sulfurovaceae: oxidize elemental sulfur  
Thioalkalispiraceae: sulfur oxidizing  
Thiohalorhabdaceae: could be a sulfur oxidizer but not much is known on the taxa so removing from analysis.  
Thiomicrospiraceae: sulfur oxidizing bacteria, but not written about in seagrass literature
Thiotrichaceae: no information on putative function, presence/absence data in seagrass so removing.  
```{r}
# read rds
all.efx <- readRDS("~/Box Sync/GSE/GrainSizeExperiment/output/sed-DA-ASVs.rds")

# filter df
efx.filtered <- filter(all.efx, Family %in% c( "Desulfatiglandaceae", "Desulfobacteraceae", "Desulfobulbaceae", "Desulfocapsaceae","Desulfosarcinaceae","Desulfovibrionaceae","Desulfuromonadaceae","Dethiosulfatibacteraceae","Sulfurimonadaceae","Sulfurovaceae","Thioalkalispiraceae","Thiomicrospiraceae"))
```  
About a third (334) of the DA ASVs (943) are associated with some kind of sulfur, methane, or nitrogen cycling in marine sediments.  
First, I want to plot by the family-genus combination name and color by order.  
```{r}
filtered.famgen.efx.plot <- ggplot(efx.filtered, aes(x = FamGen, y = x, fill = Order)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "dark grey") +
  geom_errorbar(aes(ymax = xmax, ymin = xmin), width = 0, position = position_dodge(width = 0.2)) +
  geom_point(shape = 21, size = 2.2, alpha = 0.7, position = position_dodge(width = 0.2)) +
  scale_y_continuous(limits = c(-7, 7)) +
  scale_x_discrete(limits = rev) + 
  themes +  
  #scale_fill_manual(values=fam.colors) +
  theme(legend.position = "none") +
  coord_flip() +
  labs(x = "Microbial taxa", y = "Difference in log fold abundance") +
  ggtitle("Differentially abundant taxa from BL to MP sediment") 
filtered.famgen.efx.plot
```  
# SM Fig 4 DA S Cyclers:  
Now, I want to plot only the differentially abundant taxa that are S cyclers, but color all S oxidizers white and reducers black.  

```{r}
# read rds
all.efx <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/sed-DA-ASVs.rds")

# filter df
efx.filtered <- filter(all.efx, Family %in% c( "Desulfatiglandaceae", "Desulfobacteraceae", "Desulfobulbaceae", "Desulfocapsaceae","Desulfosarcinaceae","Desulfovibrionaceae","Desulfuromonadaceae","Dethiosulfatibacteraceae","Sulfurimonadaceae","Sulfurovaceae","Thioalkalispiraceae","Thiomicrospiraceae"))

# vector of only S reduction
s.red.vec <- c("Sulfurimonadaceae","Sulfurovaceae","Thioalkalispiraceae","Thiomicrospiraceae")

# create new category labeling oxidizers and reducers
efx.filtered$category <- ifelse(efx.filtered$Family %in% s.red.vec, "S Reducer", "S Oxidizer")
  
# create colors vec
colors.vec <- c("white", "black")

annotated.efx.plot <- ggplot(efx.filtered, aes(x = Family, y = x, fill = category)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "dark grey") +
  geom_errorbar(aes(ymax = xmax, ymin = xmin), width = 0, position = position_dodge(width = 0.2)) +
  geom_point(shape = 21, size = 6, alpha = 0.7, position = position_dodge(width = 0)) +
  scale_y_continuous(limits = c(-7, 7)) +
  scale_x_discrete(limits = rev) + 
  themes +  
  scale_fill_manual(values=colors.vec) +
  coord_flip() +
  labs(x = "Microbial taxa", y = "Difference in log fold abundance") +
  theme(legend.position = "bottom", plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 20))
annotated.efx.plot

ggsave(filename = "~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/microbe-figures/DAASV-sediment-annotated.png", width=7, height =10, annotated.efx.plot)
```

Now, I know what families to isolate and visualize across populations.  

# SM Fig 6 & Fig 5, Isolated DA families by population:
Now that I know which families are of interest, I want to visualize how the relative abundance of ASVs within each of those families varies across population. This is because when I ran the NMDS plot, I saw that the variation between populations is lost when you explore at higher taxonomic levels (family versus ASV).  

I will only be exploring the families of interest from the DAA conducted at the family level.  

Load in rel.ps and subset taxa to just families of interest.  
```{r}
# load data
rel.ps <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/ps_rel.rds")

# create vec of fam of interest
fam.vec <- c( "Desulfatiglandaceae", "Desulfobacteraceae", "Desulfobulbaceae", "Desulfocapsaceae","Desulfosarcinaceae","Desulfovibrionaceae","Desulfuromonadaceae","Dethiosulfatibacteraceae","Sulfurimonadaceae","Sulfurovaceae","Thioalkalispiraceae","Thiomicrospiraceae")

# subset rel.ps based on fam.vec
DA.fam.ps <- subset_taxa(rel.ps, Family %in% fam.vec)  %>% 
            tax_glom(taxrank = 'Family') %>%
            filter_taxa( function(x) sum(x) > 0, TRUE)
```  
Manipulate ps into df for plotting.  
```{r}
# create dataframe from phyloseq object
DA.fam.df <- psmelt(DA.fam.ps)

# convert Family to a character vector from a factor to prevent R errors
DA.fam.df$Family <- as.character(DA.fam.df$Family)
```  
For each family, calculate summary statistics across treatments for each ASV.   

Summary statistics.  
```{r}
# calculate summary stats
DA.fam.sum <- summaryBy(Abundance ~ plant.origin + sed.origin + Family, data=DA.fam.df, FUN= fun)

# Create a data frame with labels for each facet
facet_labels <- data.frame(
  Family = unique(DA.fam.sum$Family),
  label = c("Pop, p = 0.28\nSed, p = 0.01\nPopxSed, p = 0.37",
            "Pop, p = 0.46\nSed, p = 0.00\nPopxSed, p = 0.74", 
            "Pop, p = 0.38\nSed, p = 0.31\nPopxSed, p = 0.67", 
            "Pop, p = 0.58\nSed, p = 0.00\nPopxSed, p = 0.20",
            "Pop, p = 0.68\nSed, p = 0.88\nPopxSed, p = 0.53",
            "Pop, p = 0.29\nSed, p = 0.00\nPopxSed, p = 0.25", 
            "Pop, p = 0.74\nSed, p = 0.01\nPopxSed, p = 0.22", 
            "Pop, p = 0.23\nSed, p = 0.00\nPopxSed, p = 0.12",
            "Pop, p = 0.12\nSed, p = 0.03\nPopxSed, p = 0.71", 
            "Pop, p = 0.41\nSed, p = 0.06\nPopxSed, p = 0.85", 
            "Pop, p = 0.21\nSed, p = 0.08\nPopxSed, p = 0.18", 
            "Pop, p = 0.73\nSed, p = 0.00\nPopxSed, p = 0.50"),
  x = c(1,2,2,2,2,2,1,2,1,1,1,2),
  y = c(Inf,Inf,Inf,Inf,Inf,Inf,Inf,Inf,Inf,Inf,Inf,Inf)
)

# Merge the labels with the summary data
DA.fam.sum <- merge(DA.fam.sum, facet_labels, by = "Family")

# Plot with annotations
DA.fam.plot <- ggplot(DA.fam.sum, aes(x = sed.origin, y = Abundance.m, group = plant.origin, color = plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = Abundance.m + Abundance.se, ymin = Abundance.m - Abundance.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels = c("COARSE", "FINE")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec, labels = c("COARSE", "FINE")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels = c("COARSE", "FINE")) +
  themes + 
  scale_x_discrete(labels = c("COARSE", "FINE")) +
  labs(y = expression("Relative Abundance"), x = expression("Sediment Origin")) +
  theme(legend.position = "bottom", plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 19.5)) +
  facet_wrap(~Family, scales = "free_y") +
  geom_text(aes(label = label, x = x, y = y, vjust = 1.2), size = 3, color = "black")
DA.fam.plot

ggsave(filename = "~/Box Sync/GSE/GrainSizeExperiment/microbe-figures/root-DATaxa.png", width=13.5, height =8, DA.fam.plot)
```
It is most likely at the ASV level that there are differences in population performance.  

Now, visualizing top two most abundant oxidizers and reducers:  
```{r}
# create vec of facets to extract
facets.vec <- c("Sulfurimonadaceae","Thiomicrospiraceae", "Desulfobacteraceae", "Desulfobulbaceae")

# filter facets
filtered.df <- DA.fam.sum %>%
  filter(Family %in% facets.vec)

# reorder family levels
filtered.df$Family <- factor(filtered.df$Family, levels = facets.vec)

# Plot with annotations
topDA.fam.plot <- ggplot(filtered.df, aes(x = sed.origin, y = Abundance.m, group = plant.origin, color = plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = Abundance.m + Abundance.se, ymin = Abundance.m - Abundance.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels = c("COARSE", "FINE")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec, labels = c("COARSE", "FINE")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels = c("COARSE", "FINE")) +
  themes + 
  scale_x_discrete(labels = c("COARSE", "FINE")) +
  labs(y = expression("Relative Abundance"), x = expression("Sediment Origin")) +
  theme(legend.position = "bottom", plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 19.5)) +
  facet_wrap(~Family, scales = "free_y") +
  geom_text(aes(label = label, x = x, y = y, vjust = 1.2), size = 3, color = "black")
topDA.fam.plot

ggsave(filename = "~/Box Sync/GSE/GrainSizeExperiment/microbe-figures/root-top-DATaxa.png", width=7.2, height =8, topDA.fam.plot)
```


# LMM on each s cycler across treatment and population:  
Need to run a separate model selection for each family.  
For Desulfatiglandaceae.  
```{r}
# isolate family
desulfatiglandaceae.df <- DA.fam.df %>% filter(Family == "Desulfatiglandaceae")

# 2 way
desulfatiglandaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=desulfatiglandaceae.df, na.action = na.exclude)
# 1 way
desulfatiglandaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=desulfatiglandaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(desulfatiglandaceae.2way, desulfatiglandaceae.1way) # p= 0.3701, drop
desulfatiglandaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=desulfatiglandaceae.df, na.action = na.exclude)
plot(desulfatiglandaceae.full)
qqnorm(resid(desulfatiglandaceae.full))
hist(resid(desulfatiglandaceae.full))
summary(desulfatiglandaceae.full) 
Anova(desulfatiglandaceae.full)
```
For Desulfobacteraceae  
```{r}
# isolate family
Desulfobacteraceae.df <- DA.fam.df %>% filter(Family == "Desulfobacteraceae")

# 2 way
Desulfobacteraceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Desulfobacteraceae.df, na.action = na.exclude)
# 1 way
Desulfobacteraceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfobacteraceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Desulfobacteraceae.2way, Desulfobacteraceae.1way) # p= 0.7357, drop
Desulfobacteraceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfobacteraceae.df, na.action = na.exclude)
plot(Desulfobacteraceae.full)
qqnorm(resid(Desulfobacteraceae.full))
hist(resid(Desulfobacteraceae.full))
summary(Desulfobacteraceae.full) 
Anova(Desulfobacteraceae.full)
```
For Desulfobulbaceae  
```{r}
# isolate family
Desulfobulbaceae.df <- DA.fam.df %>% filter(Family == "Desulfobulbaceae")

# 2 way
Desulfobulbaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Desulfobulbaceae.df, na.action = na.exclude)
# 1 way
Desulfobulbaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfobulbaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Desulfobulbaceae.2way, Desulfobulbaceae.1way) # p= 0.6651, drop
Desulfobulbaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfobulbaceae.df, na.action = na.exclude)
plot(Desulfobulbaceae.full)
qqnorm(resid(Desulfobulbaceae.full))
hist(resid(Desulfobulbaceae.full))
summary(Desulfobulbaceae.full) 
Anova(Desulfobulbaceae.full)
```
For Desulfocapsaceae  
```{r}
# isolate family
Desulfocapsaceae.df <- DA.fam.df %>% filter(Family == "Desulfocapsaceae")

# 2 way
Desulfocapsaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Desulfocapsaceae.df, na.action = na.exclude)
# 1 way
Desulfocapsaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfocapsaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Desulfocapsaceae.2way, Desulfocapsaceae.1way) # p= 0.198, drop
Desulfocapsaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfocapsaceae.df, na.action = na.exclude)
plot(Desulfocapsaceae.full)
qqnorm(resid(Desulfocapsaceae.full))
hist(resid(Desulfocapsaceae.full))
summary(Desulfocapsaceae.full) 
Anova(Desulfocapsaceae.full)
```
For Desulfosarcinaceae  
```{r}
# isolate family
Desulfosarcinaceae.df <- DA.fam.df %>% filter(Family == "Desulfosarcinaceae")

# 2 way
Desulfosarcinaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Desulfosarcinaceae.df, na.action = na.exclude)
# 1 way
Desulfosarcinaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfosarcinaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Desulfosarcinaceae.2way, Desulfosarcinaceae.1way) # p= 0.198, drop
Desulfosarcinaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfosarcinaceae.df, na.action = na.exclude)
plot(Desulfosarcinaceae.full)
qqnorm(resid(Desulfosarcinaceae.full))
hist(resid(Desulfosarcinaceae.full))
summary(Desulfosarcinaceae.full) 
Anova(Desulfosarcinaceae.full)
```
For Desulfovibrionaceae  
```{r}
# isolate family
Desulfovibrionaceae.df <- DA.fam.df %>% filter(Family == "Desulfovibrionaceae")

# 2 way
Desulfovibrionaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Desulfovibrionaceae.df, na.action = na.exclude)
# 1 way
Desulfovibrionaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfovibrionaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Desulfovibrionaceae.2way, Desulfovibrionaceae.1way) # p= 0.2489, drop
Desulfovibrionaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfovibrionaceae.df, na.action = na.exclude)
plot(Desulfovibrionaceae.full)
qqnorm(resid(Desulfovibrionaceae.full))
hist(resid(Desulfovibrionaceae.full))
summary(Desulfovibrionaceae.full) 
Anova(Desulfovibrionaceae.full)
```
For Desulfuromonadaceae  
```{r}
# isolate family
Desulfuromonadaceae.df <- DA.fam.df %>% filter(Family == "Desulfuromonadaceae")

# 2 way
Desulfuromonadaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Desulfuromonadaceae.df, na.action = na.exclude)
# 1 way
Desulfuromonadaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfuromonadaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Desulfuromonadaceae.2way, Desulfuromonadaceae.1way) # p= 0.2181, drop
Desulfuromonadaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Desulfuromonadaceae.df, na.action = na.exclude)
plot(Desulfuromonadaceae.full)
qqnorm(resid(Desulfuromonadaceae.full))
hist(resid(Desulfuromonadaceae.full))
summary(Desulfuromonadaceae.full) 
Anova(Desulfuromonadaceae.full)
```
For Dethiosulfatibacteraceae  
```{r}
# isolate family
Dethiosulfatibacteraceae.df <- DA.fam.df %>% filter(Family == "Dethiosulfatibacteraceae")

# 2 way
Dethiosulfatibacteraceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Dethiosulfatibacteraceae.df, na.action = na.exclude)
# 1 way
Dethiosulfatibacteraceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Dethiosulfatibacteraceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Dethiosulfatibacteraceae.2way, Dethiosulfatibacteraceae.1way) # p= 0.1192, drop
Dethiosulfatibacteraceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Dethiosulfatibacteraceae.df, na.action = na.exclude)
plot(Dethiosulfatibacteraceae.full)
qqnorm(resid(Dethiosulfatibacteraceae.full))
hist(resid(Dethiosulfatibacteraceae.full))
summary(Dethiosulfatibacteraceae.full) 
Anova(Dethiosulfatibacteraceae.full)
```
For Sulfurimonadaceae  
```{r}
# isolate family
Sulfurimonadaceae.df <- DA.fam.df %>% filter(Family == "Sulfurimonadaceae")

# 2 way
Sulfurimonadaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Sulfurimonadaceae.df, na.action = na.exclude)
# 1 way
Sulfurimonadaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Sulfurimonadaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Sulfurimonadaceae.2way, Sulfurimonadaceae.1way) # p= 0.7101, drop
Sulfurimonadaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Sulfurimonadaceae.df, na.action = na.exclude)
plot(Sulfurimonadaceae.full)
qqnorm(resid(Sulfurimonadaceae.full))
hist(resid(Sulfurimonadaceae.full))
summary(Sulfurimonadaceae.full) 
Anova(Sulfurimonadaceae.full)
```
For Sulfurovaceae  
```{r}
# isolate family
Sulfurovaceae.df <- DA.fam.df %>% filter(Family == "Sulfurovaceae")

# 2 way
Sulfurovaceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Sulfurovaceae.df, na.action = na.exclude)
# 1 way
Sulfurovaceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Sulfurovaceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Sulfurovaceae.2way, Sulfurovaceae.1way) # p= 0.8548, drop
Sulfurovaceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Sulfurovaceae.df, na.action = na.exclude)
plot(Sulfurovaceae.full)
qqnorm(resid(Sulfurovaceae.full))
hist(resid(Sulfurovaceae.full))
summary(Sulfurovaceae.full) 
Anova(Sulfurovaceae.full)
```
For Thioalkalispiraceae  
```{r}
# isolate family
Thioalkalispiraceae.df <- DA.fam.df %>% filter(Family == "Thioalkalispiraceae")

# 2 way
Thioalkalispiraceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Thioalkalispiraceae.df, na.action = na.exclude)
# 1 way
Thioalkalispiraceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Thioalkalispiraceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Thioalkalispiraceae.2way, Thioalkalispiraceae.1way) # p= 0.1831, drop
Thioalkalispiraceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Thioalkalispiraceae.df, na.action = na.exclude)
plot(Thioalkalispiraceae.full)
qqnorm(resid(Thioalkalispiraceae.full))
hist(resid(Thioalkalispiraceae.full))
summary(Thioalkalispiraceae.full) 
Anova(Thioalkalispiraceae.full)
```  
For Thiomicrospiraceae  
```{r}
# isolate family
Thiomicrospiraceae.df <- DA.fam.df %>% filter(Family == "Thiomicrospiraceae")

# 2 way
Thiomicrospiraceae.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=Thiomicrospiraceae.df, na.action = na.exclude)
# 1 way
Thiomicrospiraceae.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Thiomicrospiraceae.df, na.action = na.exclude)
#see if 2way can drop
anova(Thiomicrospiraceae.2way, Thiomicrospiraceae.1way) # p= 0.4999, drop
Thiomicrospiraceae.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=Thiomicrospiraceae.df, na.action = na.exclude)
plot(Thiomicrospiraceae.full)
qqnorm(resid(Thiomicrospiraceae.full))
hist(resid(Thiomicrospiraceae.full))
summary(Thiomicrospiraceae.full) 
Anova(Thiomicrospiraceae.full)
```

# SM Fig 5, Sulfur Oxidizers and Reducers across treatments:  
Now, I want to lump all sulfur oxidizers into one category and reducers as another category. Then I want to visualize their relative abundances across treatments.  
Load in relative abundance and add category for sulfur oxiders and reducers to their respective families.  
```{r}
# load data
rel.ps <- readRDS("~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/output/ps_rel.rds")

# create vec of fam of interest
fam.vec <- c( "Desulfatiglandaceae", "Desulfobacteraceae", "Desulfobulbaceae", "Desulfocapsaceae","Desulfosarcinaceae","Desulfovibrionaceae","Desulfuromonadaceae","Dethiosulfatibacteraceae","Sulfurimonadaceae","Sulfurovaceae","Thioalkalispiraceae","Thiomicrospiraceae")

# subset rel.ps based on fam.vec
DA.fam.ps <- subset_taxa(rel.ps, Family %in% fam.vec)  %>% 
            tax_glom(taxrank = 'Family') %>%
            filter_taxa( function(x) sum(x) > 0, TRUE)
```  
Manipulate ps into df for plotting.  
```{r}
# create dataframe from phyloseq object
DA.fam.df <- psmelt(DA.fam.ps)

# convert Family to a character vector from a factor to prevent R errors
DA.fam.df$Family <- as.character(DA.fam.df$Family)

# create string of families that are sulfur reducers
sreduct.vec <- c( "Desulfatiglandaceae", "Desulfobacteraceae", "Desulfobulbaceae", "Desulfocapsaceae","Desulfosarcinaceae","Desulfovibrionaceae","Desulfuromonadaceae","Dethiosulfatibacteraceae")

# create new column called category
# if Family in sreduct place Reducers in category
# else place Oxidizers in category
DA.fam.df$category <- ifelse(DA.fam.df$Family %in% sreduct.vec, "Reducers", "Oxidizers")
```   
Now, find the summary stats for each category and visualize together faceted by category.  
```{r}
# calculate summary stats
DA.cat.sum <- summaryBy(Abundance ~ plant.origin + sed.origin + category, data=DA.fam.df, FUN= fun)

# Create a data frame with labels for each facet
facet_labels <- data.frame(
  category = unique(DA.cat.sum$category),
  label = c("Pop, p = 0.59\nSed, p = 0.00\nPopxSed, p = 0.80", "Pop, p = 0.41\nSed, p = 0.63\nPopxSed, p = 0.96"),
  x = c(1,2),
  y = c(Inf, Inf)
)

# Merge the labels with the summary data
DA.cat.sum <- merge(DA.cat.sum, facet_labels, by = "category")

# plot
DA.cat.plot <- ggplot(DA.cat.sum, aes(x=sed.origin, y=Abundance.m, group=plant.origin, color=plant.origin)) +
  geom_line(linetype = 1, color = "black", position = hdodge) + 
  geom_errorbar(aes(ymax = Abundance.m + Abundance.se, ymin = Abundance.m - Abundance.se), width = 0, position = hdodge, color = "black") + 
  geom_point(aes(fill = plant.origin, shape = plant.origin), color = "black", size = 5, alpha = 0.8, position = hdodge) + 
  scale_y_continuous(expand = c(0, 0)) + 
  scale_fill_manual(name = "Population Origin", values = colors.vec, labels= c("COARSE","FINE")) +
  scale_shape_manual(name = "Population Origin", values = shapes.vec, labels= c("COARSE","FINE")) +  # Set shapes manually
  scale_color_manual(name = "Population Origin", values = colors.vec, labels= c("COARSE","FINE")) +
  themes + 
  scale_x_discrete(labels= c("COARSE","FINE")) +
  labs(y = expression("Relative Abundance"), x= expression("Sediment Origin")) +
  theme( legend.position = "bottom",plot.margin = margin(.4, .4, .4, .4, "cm")) +
  theme(text = element_text(size = 19.5)) +
  facet_wrap(~category, scales = "free_y") +
  geom_text(aes(label = label, x = x, y = y, vjust = 1.2), size = 6, color = "black")
DA.cat.plot

ggsave(filename = "~/Library/CloudStorage/Box-Box/GSE/GrainSizeExperiment/microbe-figures/root-DATaxa-category.png", width=11, height =8, DA.cat.plot)
```

# Run stats on categories of S cyclers: 

For the reducers  
```{r}
# isolate family
reducers.df <- DA.fam.df %>% filter(category == "Reducers") 

# 2 way
reducers.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=reducers.df, na.action = na.exclude)
# 1 way
reducers.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=reducers.df, na.action = na.exclude)
#see if 2way can drop
anova(reducers.2way, reducers.1way) # p= 0.8031, drop
reducers.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=reducers.df, na.action = na.exclude)
plot(reducers.full)
qqnorm(resid(reducers.full))
hist(resid(reducers.full))
summary(reducers.full) 
Anova(reducers.full)
```  
For the oxidizers  
```{r}
# isolate family
oxidizers.df <- DA.fam.df %>% filter(category == "Oxidizers") 

# 2 way
oxidizers.2way <- lmer(Abundance ~ plant.origin*sed.origin + plant.origin + sed.origin+ (1|tank.id), data=oxidizers.df, na.action = na.exclude)
# 1 way
oxidizers.1way <- lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=oxidizers.df, na.action = na.exclude)
#see if 2way can drop
anova(oxidizers.2way, oxidizers.1way) # p= 0.9617, drop
oxidizers.full <-  lmer(Abundance ~ plant.origin + sed.origin+ (1|tank.id), data=oxidizers.df, na.action = na.exclude)
plot(oxidizers.full)
qqnorm(resid(oxidizers.full))
hist(resid(oxidizers.full))
summary(oxidizers.full) 
Anova(oxidizers.full)
```